---
title: "Clustering and PCA Assignment"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## R Markdown

```{r}
install.packages("ggplot2")
library(ggplot2)

install.packages("dplyr")
library(dplyr)
```

#### Table of contents

1. [Data preparation](#Data_preparation)  
2. [Cluster analysis (agglomerative hierarchical and k-means clustering)](#Cluster_analysis)  
3. [Evaluation of cluster results](#Evaluation_cluster_results)  


#### 1. Summary 
```{r}
country_df <- read.csv("Country-data.csv")

#getting the structure of the data to see what variables are not numeric.
str(country_df)

summary(country_df)

#PCA can only be applied on numerical values. As there is a column with char value, we should either drop that from the dataset and apply PCA, or convert it to numeric values. Here, we have converted it into numeric values.  
country_df$country = as.numeric(as.factor(country_df$country))
str(country_df)

#calculating person correlation coefficient for each pair of variables
cor(country_df)
```

#### 2. Graphical analysis
```{r}
#Generating histograms on the same page
opar <- par(no.readonly = TRUE)
par(mfrow = c(3,3))
hist(country_df[, 2], main = names(country_df)[2], xlab = names(country_df)[2], xlim = c(0,200))
hist(country_df[, 3], main = names(country_df)[3], xlab = names(country_df)[3], xlim = c(0,200))
hist(country_df[, 4], main = names(country_df)[4], xlab = names(country_df)[4], xlim = c(0,20))
hist(country_df[, 5], main = names(country_df)[5], xlab = names(country_df)[5], xlim = c(0,200))
hist(country_df[, 6], main = names(country_df)[6], xlab = names(country_df)[6], xlim = c(0,150000))
hist(country_df[, 7], main = names(country_df)[7], xlab = names(country_df)[7], xlim = c(0,200))
hist(country_df[, 8], main = names(country_df)[8], xlab = names(country_df)[8], xlim = c(0,200))
hist(country_df[, 9], main = names(country_df)[9], xlab = names(country_df)[9], xlim = c(0,20))
hist(country_df[, 10], main = names(country_df)[10], xlab = names(country_df)[10], xlim = c(0,150000))
par(opar)

#Generating density plot for each variable on the same page
opar <- par(no.readonly = TRUE)
par(mfrow = c(3,3))
plot(density(country_df[, 2]), main = names(country_df)[2], xlab = names(country_df)[2])
plot(density(country_df[, 3]), main = names(country_df)[3], xlab = names(country_df)[3])
plot(density(country_df[, 4]), main = names(country_df)[4], xlab = names(country_df)[4])
plot(density(country_df[, 5]), main = names(country_df)[5], xlab = names(country_df)[5])
plot(density(country_df[, 6]), main = names(country_df)[6], xlab = names(country_df)[6])
plot(density(country_df[, 7]), main = names(country_df)[7], xlab = names(country_df)[7])
plot(density(country_df[, 8]), main = names(country_df)[8], xlab = names(country_df)[8])
plot(density(country_df[, 9]), main = names(country_df)[9], xlab = names(country_df)[9])
plot(density(country_df[, 10]), main = names(country_df)[10], xlab = names(country_df)[10])
par(opar)

#Generating boxplot for each variable on the same graph
opar <- par(no.readonly = TRUE)
boxplot(
  country_df,
  border = 'white',
  yaxt = 'n'
)
abline(h = seq(0,100,10), lty = 'dashed', lwd = 2, col = 'darkgrey')
boxplot(
  country_df,
  border = 'blue',
  yaxt = 'n',
  add = T
)
axis(2, seq(0,100,10))
par(opar)
```


#### 3. Principal Component Analysis
```{r}
pca_COUNTRY <- prcomp(country_df,scale = TRUE)
attributes(pca_)
#prcomp returns a List that has x, sdev, an rotation. x contains the PCs for drawing  graph 
#There are 167 instances and 10 variables in the dataset. Therefore, the pca should have 10 variables, each one representing one PC. However, they are ranked in order of importance and the first PC accounts for the most variation in the original data. Thus, we draw the PCA plot with merely PC1 and PC2. 
```

#### 4. Visual analysis of PCA results
### 4.1 My try
```{r}
#Creating PCA plot with x axis being PC1 and y axis being PC2 
plot(pca$x[,1], pca$x[,2])

#Calculating the variance each PC contribute to with the help of sdev which is standard deviation
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100,1)
barplot(pca.var.per, main = "Scree plot", xlab = "Principal Component", ylab = "Percent Variation")

#Getting the loading scores with the help of rotation
loading_scors <- pca$rotation[,1]
#getting the absolute values of the loading scores as we want to compare their value regardless of their sign
variable_scors <- abs(loading_scors)
#sorting the loading scores as descending
variable_scors_ranked <- sort(variable_scors, decreasing = TRUE)
#choosing top 5 variables that have the largest loading scores
top_5_factors <- names(variable_scors_ranked[1:5])
top_5_factors
#getting the top five variables with the largest loading scores with their original sign to see which variables drive the data points to left and which ones drive them to right
pca$rotation[top_5_factors,1]
```

### 4.2 From LAb_04_Tutorial: This is what I need to do for DDA as well
In addition to the scree barplot and PCA plot, scree line graph and the loading score plots are drawn. 
                   ```{r}
# perform PCA on the swiss dataset
#   note: variables are centered and scaled before analysis
pc_swiss <- prcomp(swiss, center = T, scale. = T)

# inspect the attributes of the PCA object returned by prcomp
attributes(pc_swiss)
# see value section of the help for the prcomp for more details
help(prcomp)
```

#### 4. Visual analysis of PCA results{#Visual_analysis_PCA}
```{r}
plot(pc_swiss$x[,1], pc_swiss$x[,2])

# calculate the proportion of explained variance (PEV) from the std values
pc_swiss_var <- pc_swiss$sdev^2
pc_swiss_var
pc_swiss_PEV <- pc_swiss_var / sum(pc_swiss_var)
pc_swiss_PEV

# plot the variance per PC
#   note: this can be done using the plot function on the prcomp object
plot(pc_swiss)

# plot the cumulative value of PEV for increasing number of additional PCs
#   note: add an 80% threshold line to inform the feature extraction
#     according to the plot the first 3 PCs should be selected
opar <- par(no.readonly = TRUE)
plot(
  cumsum(pc_swiss_PEV),
  ylim = c(0,1),
  xlab = 'PC',
  ylab = 'cumulative PEV',
  pch = 20,
  col = 'orange'
)
abline(h = 0.8, col = 'red', lty = 'dashed')
par(opar)

# get and inspect the loadings for each PC
#   note: loadings are reported as a rotation matrix (see lecture)
pc_swiss_loadings <- pc_swiss$rotation
pc_swiss_loadings

# plot the loadings for the first three PCs as a barplot
#   note: two vectors for colours and labels are created for convenience
#     for details on the other parameters see the help for barplot and legend
opar <- par(no.readonly = TRUE)
colvector = c('red', 'orange', 'yellow', 'green', 'cyan', 'blue')
labvector = c('PC1', 'PC2', 'PC3')
barplot(
  pc_swiss_loadings[,c(1:3)],
  beside = T,
  yaxt = 'n',
  names.arg = labvector,
  col = colvector,
  ylim = c(-1,1),
  border = 'white',
  ylab = 'loadings'
)
axis(2, seq(-1,1,0.1))
legend(
  'bottomright',
  bty = 'n',
  col = colvector,
  pch = 15,
  row.names(pc_swiss_loadings)
)
par(opar)

# generate a biplot for each pair of important PCs (and show them on the same page)
#   note: the option choices is used to select the PCs - default is 1:2
opar <- par(no.readonly = TRUE)
par(mfrow = c(2,2))
biplot(
  pc_swiss,
  scale = 0,
  col = c('grey40','orange')
)
biplot(
  pc_swiss,
  choices = c(1,3),
  scale = 0,
  col = c('grey40','orange')
)
biplot(
  pc_swiss,
  choices = c(2,3),
  scale = 0,
  col = c('grey40','orange')
)
par(opar)
```